Dark Control:\\ A Unified Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-05-09 paper/paper.tex
Multiple edits, especially mention of the synaptic epigenesis with link to policy matrix, and, of the phylogeny of connectivity with link to social interaction in human

\maketitle
\small{
    \textbf{1} Universit\'e Paris-Saclay, INRIA, Parietal Team, Saclay, France\\
    \textbf{2} Universit\'e Paris-Saclay, CEA, Neurospin, Gif-sur-Yvette, France\\
    \textbf{3} Department of Psychiatry, Psychotherapy and Psychosomatics, RWTH Aachen University, Aachen, Germany\\
    \textbf{4} Institut Pasteur, Human Genetics and Cognitive Functions Unit, Paris, France\\
    \textbf{5} CNRS UMR 3571 Genes, Synapses and Cognition, Institut Pasteur, Paris, France\\
    \textbf{6} University Paris Diderot, Sorbonne Paris Cit\'e, Paris, France\\
    \textbf{7} Centre de Bioinformatique, Biostatistique et Biologie Intégrative, Paris, France
}

@@

The present paper proposes a \textit{process model}
that tries to explain \textit{how}
the DMN may implement
continuous evaluation and prediction of the environment to guide behavior.
%
DMN function is recast in mathematical terms of
control theory and reinforcement learning based on Markov decision processes.
%
We argue that our formal account of DMN function naturally accommodates as special cases
the previously proposed cognitive accounts on
(1) predictive coding,
(2) semantic associations, and
(3) a ``sentinel'' role.
%
A process model for the neural optimization of complex behavior in the DMN
offers parsimonious explanations for
recent experimental findings in animals and humans.
\end{abstract}

% official NIPS keywords
\textbf{\\keywords}: systems biology, mind-wandering, cognitive science,
artificial intelligence,
reinforcement learning
% autonomous learning systems
% \textbf{\\Get opinion from}: Danielle Bassett, Kai Vogeley, Mohammad Shakir,
% Guillaume Dumas, Smallwood, Tenenbaum

@@

Psychologically,
the more the ongoing executed task is unknown and unpracticed,
the less stimulus-independent thoughts occur
\citep{filler1973daydreaming, teasdale1995stimulus, christoff2016mind}.
Conversely,
it is known that, the more the world is easy to predict,
the more human mental activity
becomes detached from the actual sensory environment
\citep{antrobus1966studies, pope1978regulation}.
Without requiring explicit awareness,
these ``offline'' processes may contribute
to optimizing control of the organism.
We formalize
a \textit{policy matrix} to capture the space of possible actions that
the agent can perform
on the environment given the current state. A \textit{value function}
maps environmental objects and events (i.e., states) to expected rewards.
Switching between states reduces to a sequential processing model.
Informed by outcomes of performed actions, neural computation reflected in
DMN dynamics could be constantly adapted in feedback loops
shaped by prediction error.
Such an RL account of DMN function can naturally embed human behavior
into the tension between exploitative action with immediate gains and
exploratory action with longer-term gratification.

@@

\subsection{Markov decision processes}
Model-free RL has had great success in real-world problems including robotics
\citep{ng2004,abbeel2004},
super-human performance in complex games like Atari\citep{mnih2015} and Go\citep{silver2016mastering}, etc., just to name a few.
We emphasize that the brain in general, and the DMN in particular,
is a physical system governed by the laws of
physics and can be formally described
by Markov processes at a sufficiently coarse scale.
It has indeed been previously proposed \citep{tegmark2016improved} that
any system obeying the laws of classical physics can be accurately modeled as a Markov process as long as the time
step is sufficiently short.
% , defining $s(t)$, the state at time $t$, as the position in state space.
The process has $memory$ if the next state depends not only on the current state
but also on a finite number of past states.
Rational probabilistic planning can be reformulated
as a standard memoryless Markov process by simply expanding the
definition of the state $s$ to include experience episodes of the past.

@@

\begin{mdframed}
  \textbf{ Neural correlates of the Bellman equation in the DMN:}
  ...
\end{mdframed}

@@

\begin{equation}
  \pi^{\text{hard-max}}(s) = \argmax_{a \in \mathcal A}\tilde{Q}(s, a|\theta),
  \label{eq:policy}
\end{equation}
This defines a mapping from states directly to actions.
% $a \mapsto \tilde{Q}(a,s|\theta)$ of the $Q$-value function ``frozen'' at state $s$.
For instance, a simple linear model with a kernel $\phi$ would be of the
form $\tilde{Q}(s, a|\theta) = \phi(s,a)^T\theta$, where
$\phi(s,a)$ would represent a high-level representation of the state-action pairs
$(s,a)$, as was done proposed \citep{songNIPS2016}, or a eural network as recently proposed by in the
breakthrough works
\citep{mnih2015,silver2016mastering} for playing complex games (atari, Go, etc.) at superhuman levels.
%  and Parr et al. 2003 (JMLR)].
In the DMN, the dmPFC would implement such a hard-max lookup
over the action space.
The model
  parameters $\theta$ would correspond to synaptic weights and connection strengths between
  brain regions. It is a time-varying neuronal program which dictates how to move from world states $s$ to actions $a$ via the hard-max policy \eqref{eq:policy}.
  The approximating $Q$-value function $\tilde{Q}(s, a|\theta)$ would tell the DMN the (expected) usefulness of taking an action $a$ in state $s$.
  The DMN, and in particular its dmPFC node, could then contribute to the choice, at a given state $s$, of an action $a$ which maximizes these approximate
  $Q$-values. Such a mapping from states to actions is conventionally called \textit{policy matrix}\citep{mnih2015,silver2016mastering}.
  Learning consists in starting from a given table and
  updating it during action choices,
  which take the agent to different table entries.

@@

More specifically,
environmental stimuli especially important for humans are frequently of
social nature. This is probably unsurprising
given that
the complexity of the social systems
is a likely to be a human-defining property
\citep{tomasello2009cultural}.
According to the ``social brain hypothesis'',
the human brain has especially been shaped for
forming and maintaining increasingly complex
social systems,
which allows solving ecological problems by means of social relationships
\citep{whiten1988machiavellian}.
Indeed, social topics amounted to roughly
two thirds of human everyday communication \citep{dunbar1997human},
while
mind-wandering at daytime and dreams during sleep
are rich in stories about people and
the complex relationships between them.
%
In line with this, the DMN was argued to be specialized in
continuous processing of social information as a
physiological baseline of human brain function
\citep{schilbach2008minds}. This view was later challenged by observing
analogues of the DMN in monkeys \citep{mantini2011default},
cats \citep{popa2009contrasting},
and rats \citep{lu2012rat}, three species with
social-cognitive capacities that are supposedly less advanced than in humans.
Rather than functional specificity for processing social information,
the present MDP account can parsimoniously incorporate
the dominance of social content in
human mental activity
as high value function estimates for information about humans
\citep{baker2009action, kampe2001psychology, krienen2010clan}.
The DMN may thus modulate reward processing
in the human agent in a way that prioritizes
appraisal of and action towards social contexts,
without excluding relevance of environmental cues of the physical world.
In sum,
our account on the DMN directly implies
its previously proposed ``sentinel'' function
of monitoring the environment for self-relevant information
in general and
inherently accommodates the importance of social environmental cues
as a special case.

@@

\section{Conclusion}
What brain function could be important enough
for the existence and survival of the human species
to justify constantly high energy costs?
MDPs motivate an attractive
formal account how the human association cortex
might implement multi-sensory representation and control of the environment to
optimize the organism's interaction with the world.
% by obtaining reward and avoiding pain.
This idealized process model explains
a number of previous experimental observations in the
DMN by simple but non-trivial mechanisms.
%
From the view of a Markovian sequential decision process,
human behavior unfolds by integrating action outcomes from stored past events
and extrapolation to
upcoming events for guiding action choice in the present context.
MDPs also provide a mathematical formalism how
opportunity structure in the environment can be recursively exploited
when confronted with challenging decisions.
%
This functional interpretation may be more compatible with the DMN's
poorly understood involvement across
autobiographical memory recall, problem solving,
abstract reasoning, social cognition,
as well as delay discounting and self-related prospection.
Improvement of the internal world representation
by injecting stochasticity in the recall of past
actions and the estimation of action outcomes may explain why
highly accurate memories have been disfavored in human evolution
and why human creativity might be adaptive.

>>>>

\maketitle
\small{
    \textbf{1} Universit\'e Paris-Saclay, INRIA, Parietal Team, Saclay, France\\
    \textbf{2} Universit\'e Paris-Saclay, CEA, Neurospin, Gif-sur-Yvette, France\\
    \textbf{3} Department of Psychiatry, Psychotherapy and Psychosomatics, RWTH Aachen University, Aachen, Germany\\
    \textbf{4} Institut Pasteur, Human Genetics and Cognitive Functions Unit, Paris, France\\
    \textbf{5} CNRS UMR 3571 Genes, Synapses and Cognition, Institut Pasteur, Paris, France\\
    \textbf{6} University Paris Diderot, Sorbonne Paris Cit\'e, Paris, France\\
    \textbf{7} Centre de Bioinformatique, Biostatistique et Biologie Int\'egrative, Paris, France
}

@@

The present paper proposes a \textit{process model}
that tries to explain \textit{how}
the DMN may implement
continuous evaluation and prediction of the environment to guide behavior.
%
DMN function is recast in mathematical terms of
control theory and reinforcement learning based on Markov decision processes.
%
We argue that our formal account of DMN function naturally accommodates as special cases
the previously proposed cognitive accounts on
(1) predictive coding,
(2) semantic associations, and
(3) a ``sentinel'' role.
%
Moreover, this process model for the neural optimization of complex behavior in the DMN
offers parsimonious explanations for
recent experimental findings\suggestremove{ in animals and humans}.
\end{abstract}

% official NIPS keywords
\textbf{\\keywords}: systems biology, mind-wandering, cognitive science,
artificial intelligence,
reinforcement learning
\textbf{XXX: Maybe the keywords "systems biology" and "cognitive science" are too broad. I would also remove "reinforcement learning", which is already in the title. The keywords are the occasion to enhance reachability through web and bibliography search. What about "process model", "Markov decision processes", or "Bellman equation"?}
% autonomous learning systems
% \textbf{\\Get opinion from}: Danielle Bassett, Kai Vogeley, Mohammad Shakir,
% Guillaume Dumas, Smallwood, Tenenbaum

@@

Psychologically,
the more the ongoing executed task is unknown and unpracticed,
the less stimulus-independent thoughts occur
\citep{filler1973daydreaming, teasdale1995stimulus, christoff2016mind}.
Conversely,
it is known that, the more the world is easy to predict,
the more human mental activity
becomes detached from the actual sensory environment
\citep{antrobus1966studies, pope1978regulation}.
Without requiring explicit awareness,
these ``offline'' processes may contribute
to optimizing control of the organism.
We formalize
a \textit{policy matrix} to capture the space of possible actions that
the agent can perform
on the environment given the current state. A \textit{value function}
maps environmental objects and events (i.e., states) to expected rewards.
Switching between states reduces to a sequential processing model.
Informed by outcomes of performed actions, neural computation reflected in
DMN dynamics could be constantly adapted in feedback loops
shaped by prediction error. \textbf{XXX: "could be constantly adapted in feedback loops shaped by prediction error" => "are constantly shaped by prediction error through feedback loops"}
Such an RL account of DMN function can naturally embed human behavior
into the tension between exploitative action with immediate gains and
exploratory action with longer-term gratification.

@@

\subsection{Markov decision processes}
Model-free RL has had great success in real-world problems including robotics
\citep{ng2004,abbeel2004},
super-human performance in complex video games \citep{mnih2015} and Go\citep{silver2016mastering}, etc., just to name a few.
We emphasize that the brain in general, and the DMN in particular,
is a physical system governed by the laws of
physics and can be formally described
by Markov processes at a sufficiently coarse scale.
It has indeed been previously proposed \citep{tegmark2016improved} that
any system obeying the laws of classical physics can be accurately modeled as a Markov process as long as the time
step is sufficiently short.
% , defining $s(t)$, the state at time $t$, as the position in state space.
The process has $memory$ if the next state depends not only on the current state
but also on a finite number of past states.
Rational probabilistic planning can be reformulated
as a standard memoryless Markov process by simply expanding the
definition of the state $s$ to include experience episodes of the past.

@@

\begin{mdframed}
  \textbf{ Neural correlates of the Bellman equation in the DMN:}
  ...
  \textbf{XXX Guillaume: I agree that this would be BIG!}
\end{mdframed}

@@

\begin{equation}
  \pi^{\text{hard-max}}(s) = \argmax_{a \in \mathcal A}\tilde{Q}(s, a|\theta),
  \label{eq:policy}
\end{equation}
This defines a mapping from states directly to actions.
% $a \mapsto \tilde{Q}(a,s|\theta)$ of the $Q$-value function ``frozen'' at state $s$.
For instance, a simple linear model with a kernel $\phi$ would be of the
form $\tilde{Q}(s, a|\theta) = \phi(s,a)^T\theta$, where
$\phi(s,a)$ would represent a high-level representation of the state-action pairs
$(s,a)$, as was done proposed \citep{songNIPS2016}, or a eural network as recently proposed by in the
breakthrough works
\citep{mnih2015,silver2016mastering} for playing complex games (atari, Go, etc.) at superhuman levels.
%  and Parr et al. 2003 (JMLR)].
In the DMN, the dmPFC would implement such a hard-max lookup
over the action space.
The model
  parameters $\theta$ would correspond to synaptic weights and connection strengths between
  brain regions. It is a time-varying neuronal program which dictates how to move from world states $s$ to actions $a$ via the hard-max policy \eqref{eq:policy}.
  The approximating $Q$-value function $\tilde{Q}(s, a|\theta)$ would tell the DMN the (expected) usefulness of taking an action $a$ in state $s$.
  The DMN, and in particular its dmPFC node, could then contribute to the choice, at a given state $s$, of an action $a$ which maximizes these approximate
  $Q$-values. Such a mapping from states to actions is conventionally called \textit{policy matrix}\citep{mnih2015,silver2016mastering}.
  Learning consists in starting from a given table and
  updating it during action choices,
  which take the agent to different table entries. \suggestadd{An additional layer of learning concerns the addition of new entries in the table (i.e. state and action). The framing of the policy matrix is not described here but could be supported by the formation of new attractor in the neural dynamics through synaptic epigenesis \citep{gisiger_acquisition_2005}.}

@@

More specifically,
environmental stimuli especially important for humans are frequently of
social nature. This is probably unsurprising
given that
the complexity of the social systems
is a likely to be a human-defining property
\citep{tomasello2009cultural}.
According to the ``social brain hypothesis'',
the human brain has especially been shaped for
forming and maintaining increasingly complex
social systems,
which allows solving ecological problems by means of social relationships
\citep{whiten1988machiavellian}.
Indeed, social topics amounted to roughly
two thirds of human everyday communication \citep{dunbar1997human},
while
mind-wandering at daytime and dreams during sleep
are rich in stories about people and
the complex relationships between them.
%
In line with this, the DMN was argued to be specialized in
continuous processing of social information as a
physiological baseline of human brain function
\citep{schilbach2008minds}. This view was later challenged by observing
analogues of the DMN in monkeys \citep{mantini2011default},
cats \citep{popa2009contrasting},
and rats \citep{lu2012rat}, three species with
social-cognitive capacities that are supposedly less advanced than in humans.

\suggestadd{The principal connectivity gradient in the cortex appears to be greatly expanded in humans compare to macaques, suggesting a phylogenetically conserved axis of cortical expansion with the DMN emerging at the extreme-end in humans \citep{margulies_situating_2016}. Neurocomputational models of dyadic whole brain demonstrated how the human connectivity topology, on top of facilitating processing at the intra-individual level, can explain our propensity to coordinate through sensorimotor loops with others at the inter-individual level \citep{dumas_anatomical_2012}.}

Rather than functional specificity for processing social information,
the present MDP account can parsimoniously incorporate
the dominance of social content in
human mental activity
as high value function estimates for information about humans
\citep{baker2009action, kampe2001psychology, krienen2010clan}.
The DMN may thus modulate reward processing
in the human agent in a way that prioritizes
appraisal of and action towards social contexts,
without excluding relevance of environmental cues of the physical world.
In sum,
our account on the DMN directly implies
its previously proposed ``sentinel'' function
of monitoring the environment for self-relevant information
in general and
inherently accommodates the importance of social environmental cues
as a special case.

@@

\section{Conclusion}
Which brain function could be important enough
for the existence and survival of the human species
to justify constantly high energy costs?
MDPs motivate an attractive
formal account how the human association cortex
might implement multi-sensory representation and control of the environment to
optimize the organism's interaction with the world.
% by obtaining reward and avoiding pain.
This idealized process model explains
a number of previous experimental observations in the
DMN by simple but non-trivial mechanisms.
%
From the view of a Markovian sequential decision process,
human behavior unfolds by integrating action outcomes from stored past events
and extrapolation to
upcoming events for guiding action choice in the present context.
MDPs also provide a mathematical formalism how
opportunity structure in the environment can be recursively exploited
when confronted with challenging decisions.
%
This functional interpretation may be more compatible with the DMN's
poorly understood involvement across
autobiographical memory recall, problem solving,
abstract reasoning, social cognition,
as well as delay discounting and self-related prospection.
Improvement of the internal world representation
by injecting stochasticity in the recall of past
actions and the estimation of action outcomes may explain why
highly accurate memories have been disfavored in human evolution
and why human creativity might be adaptive.
